---
title: "Regression in R"
author: "Riyanshi Bohra"
format: html
editor: visual
---

# Setting Up

```{r}
# Install necessary packages
if (!require(pacman))
  install.packages("pacman")

pacman::p_load(tidymodels,
               tidyverse,
               ranger,
               randomForest,
               glmnet,
               gridExtra)

```

```{r}
# Global theme
theme_set(theme_bw() + theme(legend.position = "top"))

```

# Synthetic Data Generation

```{r}
# This block sets parameters for the analysis.
seed <- 1                   # Set the random seed for reproducibility
numInstances <- 200         # Define the total number of instances in the dataset

# Set seed
set.seed(seed)

# Generating data
X <- matrix(runif(numInstances), ncol=1)
y_true <- -3*X + 1 
y <- y_true + matrix(rnorm(numInstances), ncol=1)

```

```{r}
# Plotting the generated data to visualize the linear relationship

ggplot() +
  geom_point(aes(x=X, y=y), color="black") +
  geom_line(aes(x=X, y=y_true), color="blue", linewidth=1) +
  ggtitle('True function: y = -3X + 1') +
  xlab('X') +
  ylab('y')           # Create a scatter plot of the generated data

# Interpretation:
# Points represent the generated data, and the blue line shows the true underlying linear relationship without the noise
```

# Multiple Linear Regression

## 1) Split Input Data into Training and Test Sets

```{r}
# This section splits the data into training and test sets
numTrain <- 20   # number of training instances
numTest <- numInstances - numTrain

set.seed(123) # For reproducibility
```

```{r}
# Combine X and y into a tibble for easier data manipulation
data <- tibble(X = X, y = y)

split_obj <- initial_split(data, prop = numTrain/numInstances)
```

```{r}
# Split the dataset into training and testing sets
train_data <- training(split_obj)
test_data <- testing(split_obj)

# Extract training and testing data from the split object.
X_train <- train_data$X
y_train <- train_data$y

X_test <- test_data$X
y_test <- test_data$y
```

## 2) Fit Regression Model to Training Set

```{r}
# Setting up a linear regression model using the 'lm' engine
lin_reg_spec <- linear_reg() |> 
  set_engine("lm")

# Fit the linear regression model to the training data
lin_reg_fit <- lin_reg_spec |> 
  fit(y ~ X, data = train_data)
```

## 3) Apply Model to the Test Set

```{r}
# Predict response values for the test set using the fitted model
y_pred_test <- predict(lin_reg_fit, new_data = test_data) |>
  pull(.pred)
```

## 4) Evaluate Model Performance on Test Set

```{r}
# Create a scatter plot to compare true and predicted values of the response variable for the test set
ggplot() + 
  geom_point(aes(x = as.vector(y_test), y = y_pred_test), color = 'black') +
  ggtitle('Comparing true and predicted values for test set') +
  xlab('True values for y') +
  ylab('Predicted values for y')
```

```{r}
# Organize test data and predictions for model evaluation
eval_data <- tibble(
  truth = as.vector(y_test),
  estimate = y_pred_test
)

# Calculate and display the Root Mean Squared Error (RMSE) of the model
rmse_value <- rmse(data = eval_data, truth = truth, estimate = estimate)
r2_value <- rsq(eval_data, truth = truth, estimate = estimate)
```

```{r}
# Output the RMSE and R-squared value to assess the model's goodness of fit
cat("Root mean squared error =", sprintf("%.4f", rmse_value$.estimate), "\n")
cat('R-squared =', sprintf("%.4f", r2_value$.estimate), "\n")
```

## 5) Postprocessing

```{r}
# Extract and display the slope coefficient from the linear regression model
coef_values <- coef(lin_reg_fit$fit) 
slope <- coef_values["X"]
intercept <- coef_values["(Intercept)"]


```

```{r}
# Output the intercept of the linear regression model
cat("Slope =", slope, "\n")

cat("Intercept =", intercept, "\n")
```

```{r}
### Step 4: Postprocessing

# Plot outputs
ggplot() +
  geom_point(aes(x = as.vector(X_test), y = as.vector(y_test)), color = 'black') +
  geom_line(aes(x = as.vector(X_test), y = y_pred_test), color = 'blue', linewidth = 1) +
  ggtitle(sprintf('Predicted Function: y = %.2fX + %.2f', slope, intercept)) +
  xlab('X') +
  ylab('y')
```

# Effect of Correlated Attributes

## Create correlated attributes

```{r}
# Generate the variables
set.seed(1)
X2 <- 0.5 * X + rnorm(numInstances, mean=0, sd=0.04)
X3 <- 0.5 * X2 + rnorm(numInstances, mean=0, sd=0.01)
X4 <- 0.5 * X3 + rnorm(numInstances, mean=0, sd=0.01)
X5 <- 0.5 * X4 + rnorm(numInstances, mean=0, sd=0.01)

# Create plots
plot1 <- ggplot() +
  geom_point(aes(X, X2), color='black') +
  xlab('X') + ylab('X2') +
  ggtitle(sprintf("Correlation between X and X2 = %.4f", cor(X[-c((numInstances-numTest+1):numInstances)], X2[-c((numInstances-numTest+1):numInstances)])))

plot2 <- ggplot() +
  geom_point(aes(X2, X3), color='black') +
  xlab('X2') + ylab('X3') +
  ggtitle(sprintf("Correlation between X2 and X3 = %.4f", cor(X2[-c((numInstances-numTest+1):numInstances)], X3[-c((numInstances-numTest+1):numInstances)])))

plot3 <- ggplot() +
  geom_point(aes(X3, X4), color='black') +
  xlab('X3') + ylab('X4') +
  ggtitle(sprintf("Correlation between X3 and X4 = %.4f", cor(X3[-c((numInstances-numTest+1):numInstances)], X4[-c((numInstances-numTest+1):numInstances)])))

plot4 <- ggplot() +
  geom_point(aes(X4, X5), color='black') +
  xlab('X4') + ylab('X5') +
  ggtitle(sprintf("Correlation between X4 and X5 = %.4f", cor(X4[-c((numInstances-numTest+1):numInstances)], X5[-c((numInstances-numTest+1):numInstances)])))

# Combine plots into a 2x2 grid
grid.arrange(plot1, plot2, plot3, plot4, ncol=2)
```

```{r}
# Split data into training and testing sets
train_indices <- 1:(numInstances - numTest)
test_indices <- (numInstances - numTest + 1):numInstances

# Create combined training and testing sets
X_train2 <- cbind(X[train_indices], X2[train_indices])
X_test2 <- cbind(X[test_indices], X2[test_indices])

X_train3 <- cbind(X[train_indices], X2[train_indices], X3[train_indices])
X_test3 <- cbind(X[test_indices], X2[test_indices], X3[test_indices])

X_train4 <- cbind(X[train_indices], X2[train_indices], X3[train_indices], X4[train_indices])
X_test4 <- cbind(X[test_indices], X2[test_indices], X3[test_indices], X4[test_indices])

X_train5 <- cbind(X[train_indices], X2[train_indices], X3[train_indices], X4[train_indices], X5[train_indices])
X_test5 <- cbind(X[test_indices], X2[test_indices], X3[test_indices], X4[test_indices], X5[test_indices])
```

```{r}
# Convert matrices to tibbles for training
train_data2 <- tibble(X1 = X_train2[,1], X2 = X_train2[,2], y = y_train)
train_data3 <- tibble(X1 = X_train3[,1], X2 = X_train3[,2], X3 = X_train3[,3], y = y_train)
train_data4 <- tibble(X1 = X_train4[,1], X2 = X_train4[,2], X3 = X_train4[,3], X4 = X_train4[,4], y = y_train)
train_data5 <- tibble(X1 = X_train5[,1], X2 = X_train5[,2], X3 = X_train5[,3], X4 = X_train5[,4], X5 = X_train5[,5], y = y_train)

# Train models
regr2_spec <- linear_reg() %>% set_engine("lm")
regr2_fit <- regr2_spec %>% fit(y ~ X1 + X2, data = train_data2)

regr3_spec <- linear_reg() %>% set_engine("lm")
regr3_fit <- regr3_spec %>% fit(y ~ X1 + X2 + X3, data = train_data3)

regr4_spec <- linear_reg() %>% set_engine("lm")
regr4_fit <- regr4_spec %>% fit(y ~ X1 + X2 + X3 + X4, data = train_data4)

regr5_spec <- linear_reg() %>% set_engine("lm")
regr5_fit <- regr5_spec %>% fit(y ~ X1 + X2 + X3 + X4 + X5, data = train_data5)
```

```{r}
# Convert matrices to data.frames for predictions
new_train_data2 <- setNames(as.data.frame(X_train2), c("X1", "X2"))
new_test_data2 <- setNames(as.data.frame(X_test2), c("X1", "X2"))

new_train_data3 <- setNames(as.data.frame(X_train3), c("X1", "X2", "X3"))
new_test_data3 <- setNames(as.data.frame(X_test3), c("X1", "X2", "X3"))

new_train_data4 <- setNames(as.data.frame(X_train4), c("X1", "X2", "X3", "X4"))
new_test_data4 <- setNames(as.data.frame(X_test4), c("X1", "X2", "X3", "X4"))

new_train_data5 <- setNames(as.data.frame(X_train5), c("X1", "X2", "X3", "X4", "X5"))
new_test_data5 <- setNames(as.data.frame(X_test5), c("X1", "X2", "X3", "X4", "X5"))

# Predictions
y_pred_train2 <- predict(regr2_fit, new_data = new_train_data2)
y_pred_test2 <- predict(regr2_fit, new_data = new_test_data2)

y_pred_train3 <- predict(regr3_fit, new_data = new_train_data3)
y_pred_test3 <- predict(regr3_fit, new_data = new_test_data3)

y_pred_train4 <- predict(regr4_fit, new_data = new_train_data4)
y_pred_test4 <- predict(regr4_fit, new_data = new_test_data4)

y_pred_train5 <- predict(regr5_fit, new_data = new_train_data5)
y_pred_test5 <- predict(regr5_fit, new_data = new_test_data5)
```

```{r}
# Extract coefficients and intercepts
get_coef <- function(model) {
  coef <- coefficients(model$fit)
  coef
}

# Calculate RMSE
calculate_rmse <- function(actual, predicted) {
  rmse <- sqrt(mean((actual - predicted)^2))
  rmse
}

results <- tibble(
  Model = c(sprintf("%.2f X + %.2f", get_coef(regr2_fit)['X1'], get_coef(regr2_fit)['(Intercept)']),
            sprintf("%.2f X + %.2f X2 + %.2f", get_coef(regr3_fit)['X1'], get_coef(regr3_fit)['X2'], get_coef(regr3_fit)['(Intercept)']),
            sprintf("%.2f X + %.2f X2 + %.2f X3 + %.2f", get_coef(regr4_fit)['X1'], get_coef(regr4_fit)['X2'], get_coef(regr4_fit)['X3'], get_coef(regr4_fit)['(Intercept)']),
            sprintf("%.2f X + %.2f X2 + %.2f X3 + %.2f X4 + %.2f", get_coef(regr5_fit)['X1'], get_coef(regr5_fit)['X2'], get_coef(regr5_fit)['X3'], get_coef(regr5_fit)['X4'], get_coef(regr5_fit)['(Intercept)'])),
  
  Train_error = c(calculate_rmse(y_train, y_pred_train2$.pred),
                  calculate_rmse(y_train, y_pred_train3$.pred),
                  calculate_rmse(y_train, y_pred_train4$.pred),
                  calculate_rmse(y_train, y_pred_train5$.pred)),
  
  Test_error = c(calculate_rmse(y_test, y_pred_test2$.pred),
                 calculate_rmse(y_test, y_pred_test3$.pred),
                 calculate_rmse(y_test, y_pred_test4$.pred),
                 calculate_rmse(y_test, y_pred_test5$.pred)),
  
  Sum_of_Absolute_Weights = c(sum(abs(get_coef(regr2_fit))),
                              sum(abs(get_coef(regr3_fit))),
                              sum(abs(get_coef(regr4_fit))),
                              sum(abs(get_coef(regr5_fit))))
)

# Plotting
ggplot(results, aes(x = Sum_of_Absolute_Weights)) +
  geom_line(aes(y = Train_error, color = "Train error"), linetype = "solid") +
  geom_line(aes(y = Test_error, color = "Test error"), linetype = "dashed") +
  labs(x = "Sum of Absolute Weights", y = "Error rate") +
  theme_minimal()
```

```{r}
results
```

# Ridge Regression

```{r}
# Convert to data frame
train_data <- tibble(y = y_train, X_train5)
test_data <- tibble(y = y_test, X_test5)

# Set up a Ridge regression model specification
ridge_spec <- linear_reg(penalty = 0.4, mixture = 1) %>% 
  set_engine("glmnet")

# Fit the model
ridge_fit <- ridge_spec %>% 
  fit(y ~ ., data = train_data)

# Make predictions
y_pred_train_ridge <- predict(ridge_fit, new_data = train_data)$.pred
y_pred_test_ridge <- predict(ridge_fit, new_data = test_data)$.pred


# Make predictions
y_pred_train_ridge <- predict(ridge_fit, new_data = train_data)$.pred
y_pred_test_ridge <- predict(ridge_fit, new_data = train_data)$.pred

# Calculate RMSE
calculate_rmse <- function(actual, predicted) {
  rmse <- sqrt(mean((actual - predicted)^2))
  rmse
}

# Extract coefficients
ridge_coef <- coefficients(ridge_fit$fit)

model6 <- sprintf("%.2f X + %.2f X2 + %.2f X3 + %.2f X4 + %.2f X5 + %.2f", 
                 ridge_coef[2], ridge_coef[3], ridge_coef[4], 
                 ridge_coef[5], ridge_coef[6], ridge_coef[1])

values6 <- tibble(
  Model = model6,
  Train_error = calculate_rmse(y_train, y_pred_train_ridge),
  Test_error = calculate_rmse(y_test, y_pred_test_ridge),
  Sum_of_Absolute_Weights = sum(abs(ridge_coef))
)

# Combining the results
final_results <- bind_rows(results, values6)

final_results
```

# Lasso Regression

```{r}
# Define the lasso specification
lasso_spec <- linear_reg(penalty = 0.02, mixture = 1) %>% 
  set_engine("glmnet")

# Ensure the data is combined correctly
train_data <- tibble(y = y_train, X1 = X_train5[,1], X2 = X_train5[,2], 
                     X3 = X_train5[,3], X4 = X_train5[,4], X5 = X_train5[,5])

# Fit the model
lasso_fit <- lasso_spec %>%
  fit(y ~ ., data = train_data)

# Extract coefficients
lasso_coefs <- lasso_fit$fit$beta[,1]

# Predictions
y_pred_train_lasso <- predict(lasso_fit, new_data = train_data)$.pred
y_pred_test_lasso <- predict(lasso_fit, new_data = tibble(X1 = X_test5[,1], X2 = X_test5[,2], 
                                                          X3 = X_test5[,3], X4 = X_test5[,4], X5 = X_test5[,5]))$.pred

# Create the model string
model7 <- sprintf("%.2f X + %.2f X2 + %.2f X3 + %.2f X4 + %.2f X5 + %.2f", 
                  lasso_coefs[2], lasso_coefs[3], lasso_coefs[4], 
                  lasso_coefs[5], lasso_coefs[6], lasso_fit$fit$a0[1])

values7 <- c(model7, 
             sqrt(mean((y_train - y_pred_train_lasso)^2)),
             sqrt(mean((y_test - y_pred_test_lasso)^2)),
             sum(abs(lasso_coefs[-1])) + abs(lasso_fit$fit$a0[1]))

# Make the results tibble
lasso_results <- tibble(Model = "Lasso",
                        `Train error` = values7[2], 
                        `Test error` = values7[3], 
                        `Sum of Absolute Weights` = values7[4])

lasso_results
```

# Hyperparameter Selection via Cross-Validation

```{r}
# Combine training data
y_train <- as.vector(y_train)

train_data <- tibble(y = y_train, X1 = X_train5[,1], X2 = X_train5[,2], 
                     X3 = X_train5[,3], X4 = X_train5[,4], X5 = X_train5[,5])

# Define recipe
recipe_obj <- recipe(y ~ ., data = train_data) %>%
  step_normalize(all_predictors()) |>
  prep()

# Define the ridge specification
ridge_spec <- linear_reg(penalty = tune(), mixture = 0) %>% 
  set_engine("glmnet")

# Ridge workflow
ridge_wf <- workflow() |>
  add_model(ridge_spec) |>
  add_recipe(recipe_obj)

# Grid of alphas
alphas <- tibble(penalty = c(0.2, 0.4, 0.6, 0.8, 1.0))

# Tune
tune_results <- 
  ridge_wf |>
  tune_grid(
  resamples = bootstraps(train_data, times = 5),
  grid = alphas
)


# Extract best parameters
best_params <- tune_results %>% select_best("rmse")

# Refit the model
ridge_fit <- ridge_spec %>%
  finalize_model(best_params) %>%
  fit(y ~ ., data = train_data)

# Extract coefficients
ridge_coefs <- ridge_fit$fit$beta[,1]

# Predictions
y_pred_train_ridge <- predict(ridge_fit, new_data = train_data)$.pred
y_pred_test_ridge <- predict(ridge_fit, new_data = tibble(X1 = X_test5[,1], X2 = X_test5[,2], 
                                                          X3 = X_test5[,3], X4 = X_test5[,4], X5 = X_test5[,5]))$.pred

# Create the model string
model6 <- sprintf("%.2f X + %.2f X2 + %.2f X3 + %.2f X4 + %.2f X5 + %.2f", 
                  ridge_coefs[2], ridge_coefs[3], ridge_coefs[4], 
                  ridge_coefs[5], ridge_coefs[6], ridge_fit$fit$a0[1])

values6 <- c(model6, 
             sqrt(mean((y_train - y_pred_train_ridge)^2)),
             sqrt(mean((y_test - y_pred_test_ridge)^2)),
             sum(abs(ridge_coefs[-1])) + abs(ridge_fit$fit$a0[1]))

# Make the results tibble
ridge_results <- tibble(Model = "RidgeCV",
                        `Train error` = values6[2], 
                        `Test error` = values6[3], 
                        `Sum of Absolute Weights` = values6[4])

cat("Selected alpha =", best_params$penalty, "\n")
```

```{r}
all_results <- bind_rows(results, ridge_results)
all_results
```

```{r}
set.seed(1234)

# Ensure y_train is a vector
y_train <- as.vector(y_train)

# Combine training data
train_data <- tibble(y = y_train, X1 = X_train5[,1], X2 = X_train5[,2], 
                     X3 = X_train5[,3], X4 = X_train5[,4], X5 = X_train5[,5])

# Define recipe
recipe_obj_lasso <- recipe(y ~ ., data = train_data) %>%
  step_normalize(all_predictors()) |>
  prep()

# Define the lasso specification
lasso_spec <- linear_reg(penalty = tune(), mixture = 1) %>% 
  set_engine("glmnet")

# Lasso workflow
lasso_wf <- workflow() |>
  add_recipe(recipe_obj_lasso)

# Lasso fit
lasso_fit <- lasso_wf |>
  add_model(lasso_spec) |>
  fit(data = train_data)

# Grid of alphas for Lasso
lambda_grid <- grid_regular(penalty(), levels = 50)

# Tune
tune_results_lasso <- 
  tune_grid(lasso_wf |> add_model(lasso_spec),
  resamples = bootstraps(train_data, times = 5),
  grid = lambda_grid
)

# Extract best parameters for Lasso
best_params_lasso <- tune_results_lasso %>% select_best("rmse")

# Refit the model using Lasso
lasso_fit <- lasso_spec %>%
  finalize_model(best_params_lasso) %>%
  fit(y ~ ., data = train_data)

# Extract coefficients
lasso_coefs <- lasso_fit$fit$beta[,1]

# Predictions using Lasso
y_pred_train_lasso <- predict(lasso_fit, new_data = train_data)$.pred
y_pred_test_lasso <- predict(lasso_fit, new_data = tibble(X1 = X_test5[,1], X2 = X_test5[,2], 
                                                          X3 = X_test5[,3], X4 = X_test5[,4], X5 = X_test5[,5]))$.pred

# Create the model string for Lasso
model7 <- sprintf("%.2f X + %.2f X2 + %.2f X3 + %.2f X4 + %.2f X5 + %.2f", 
                  lasso_coefs[2], lasso_coefs[3], lasso_coefs[4], 
                  lasso_coefs[5], lasso_coefs[6], lasso_fit$fit$a0[1])

values7 <- c(model7, 
             sqrt(mean((y_train - y_pred_train_lasso)^2)),
             sqrt(mean((y_test - y_pred_test_lasso)^2)),
             sum(abs(lasso_coefs[-1])) + abs(lasso_fit$fit$a0[1]))

# Make the results tibble for Lasso
lasso_results <- tibble(Model = "LassoCV",
                        `Train error` = values7[2], 
                        `Test error` = values7[3], 
                        `Sum of Absolute Weights` = values7[4])

cat("Selected alpha for Lasso =", best_params_lasso$penalty, "\n")
```

```{r}
lasso_results
```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```
